# Wrangle Rows
The previous module demonstrated methods for wrangling columns (i.e., variables) in a data frame. In this module, methods to wrangle rows, which with tidy data are observations, in a data frame are introduced. The primary method of interest here is selecting a smaller subset of rows (i.e., filtering) for further analysis.

::: {.tip data-latex=""}
Manipulating **rows** means you are manipulating **observations**.
:::

Again the descriptive examples below will use the `bears` data frame from Section \@ref(csv-files).

```{r}
bears <- read_csv(file.path("data","Bears.csv"))
bears
```

The code here will also use the pipe operator, `%>%`, to again help you become more comfortable with its use.

&nbsp;

## Selecting Specific Rows
Specific rows may be selected or omitted from a data frame using `slice()`. Below are four simple examples.

```{r}
bears %>% slice(1)                # First row
bears %>% slice(c(1,3,5))         # First, third, and fifth rows
bears %>% slice(-1)               # All but the first row
bears %>% slice(-c(1,3,5))        # All but the first, third, and fifth rows
```

Rows from the beginning (i.e., the "head") or end (i.e., the "tail") of the data frame may also be selected with `slice_head()` or `slice_tail()`, respectively. You may select a certain number of rows with `n=` or an approximate proportion of rows with `prop=`. Below are four examples.

```{r}
bears %>% slice_head(n=3)         # First three rows
bears %>% slice_head(prop=0.33)   # Approx. first 33% of rows
bears %>% slice_tail(n=3)         # Last three rows
bears %>% slice_tail(prop=0.33)   # Approx. last 33% of rows
```

Finally a random sample of rows from the data frame may be selected with `slice_sample()`, again either using `n=` or `prop=`.

```{r}
bears %>% slice_sample(n=3)       # 3 random rows
bears %>% slice_sample(prop=0.33) # Random approx. 33% rows. 
```

::: {.tip data-latex=""}
Use the `slice()` family of functions to select specific (or random) rows from a data frame.
:::

&nbsp;

## Filtering Rows
Observations or rows can be selected from a data frame with `filter()`. The directive arguments to `filter()` are conditional expressions describing which observations from the data frame to maintain. Common operators used in these conditional expressions are in Table \@ref(tab:filterComps).

&nbsp;

```{r filterComps, echo=FALSE}
data.frame(
  "Comparison Operator"=
    c('`var==value`','`var!=value`','`var %in% c(value1,value2)`',
      '`var`>`value`','`var`>=`value`','`var`<`value`','`var`<=`value`',
      'condition1,condition2','condition1 | condition2'),
  "Rows Returned from Original Data Frame"=
    c('All rows where `var` **IS equal** to `value`',
      'All rows where `var` **is NOT equal** to `value`',
      'All rows where `var` **IS IN** (or one of the) vector of `value`s^[`value` should be a character, factor, or integer.]',
      'All rows where `var` is **greater than** `value`^[`value` must be numeric.]',
      'All rows where `var` is **greater than or equal to** `value`^[`value` must be numeric.]',
      'All rows where `var` is **less than** `value`^[`value` must be numeric.]',
      'All rows where `var` is **less than or equal to** `value`^[`value` must be numeric.]',
      'All rows where **BOTH** conditions are true',
      'All rows where **ONE or BOTH** conditions are true^[Note that this "or" operator is a "vertical line"" which is typed with the shift-backslash key.]'),check.names=FALSE) %>%
  knitr::kable(booktabs=TRUE,caption="Comparison operators used in `filterD()` and their results. Note that `var` generically represents a variable in the original data frame and `value` is a generic value or level. Both `var` and `val` would be replaced with specific items (see examples in main text).") %>%
  kableExtra::kable_classic("hover",full_width=FALSE,html_font=khfont) %>%
  kableExtra::row_spec(0,bold=TRUE)%>%
  kableExtra::column_spec(1,width="2.5in")
```

&nbsp;

The following are examples of new data frames created from `bears`. The name of the new data frame (i.e., object left of the assignment operator) is `tmp` (for temporary) in each example below because there is no plan to use these data frames further.

* Only observations from *Bayfield* county.

```{r}
tmp <- bears %>% filter(loc=="Bayfield")
tmp
```

* Observations from both *Bayfield* and *Ashland* counties.

```{r}
tmp <- bears %>% filter(loc %in% c("Bayfield","Ashland"))
tmp
```

* Observations *NOT* from *Bayfield* county.

```{r}
tmp <- bears %>% filter(loc != "Bayfield")
tmp
```

* Observations with a weight greater than 100 kg.

```{r}
tmp <- bears %>% filter(weight.kg>100)
tmp
```

* Observations from *Douglas* County that weighed at least 110 kg.

```{r}
tmp <- bears %>% filter(loc=="Douglas",weight.kg>=110)
tmp
```

The last example above illustrates that multiple conditional expressions in `filter()` are combined as an "and" operator such that both conditions must be true.

::: {.tip data-latex=""}
Use `filter()` to select rows from a data frame that match a logical condition.
:::

&nbsp;

It is good practice to examine a data frame after filtering to be sure that the new data frame contains the observations that you want. The data frames above are so small that you can simply and easily examine the entire data frame. However, this will not be the case with more realistic larger data frames. Thus, I suggest the following methods for "checking your filtering."

* Simply display the data frame or the structure of the data frame to identify any obvious issues. For example, the code below should return all bears from Douglas County with a weight greater than 150 kg. Showing the data frame or the structure of the data frame both show that this data frame contains no data.

```{r}
tmp <- bears %>% filter(loc=="Douglas",weight.kg>=150)
tmp
str(tmp,give.attr=FALSE)
```

* If you filter with respect to a categorical variable then use `unique()` with that categorical variable to examine its levels. For example, the filter below is expected to return observations for just Ashland and Bayfield counties. The use of `unique()` supports that this is what is returned.

```{r}
tmp <- bears %>% filter(loc %in% c("Bayfield","Ashland"))
unique(tmp$loc)
```

* If you filter with respect to a quantitative variable then use `summary()` with that quantitative variable to examine its summary statistics. For example the fitler below is expected to return observations for lengths between 130 and 145 cm. The minimum and maximum values in the `summary()` results support that is what is returned.

```{r}
tmp <- bears %>% filter(length.cm>130,length.cm<145)
summary(tmp$length.cm)
```

::: {.tip data-latex=''}
Examine the new data frame after filtering to ensure that it contains the observations you intended.
:::

&nbsp;

## Arranging Rows
The `arrange()` function is used to sort rows based on values in one or more variables.^[Some examples of `arrange()` are in Section \@ref(columns-examples).] The default is ascending order. To sort in descending order then wrap the variable name in `desc()`. If more than one variable is given then the rows are first sorted based on the first variable and then ties in the first variable are sorted based on the second variable. Examples of sorting are shown below.

* Alphabetically sort bears by location name.

```{r}
bears <- bears %>% arrange(loc)
bears
```

* Sort bears from heaviest to lightest.

```{r}
bears <- bears %>% arrange(desc(weight.kg))
bears
```

* Sort bears from heaviest to lightest *within* each location.

```{r}
bears <- bears %>% arrange(loc,desc(weight.kg))
bears
```

* sort bears by size, first by length and then by weight.

```{r}
bears <- bears %>% arrange(length.cm,weight.kg)
bears
```

::: {.tip data-latex=""}
Use `arrange()` to sort rows in a data frame by the value(s) of variable(s).
:::

&nbsp;

## Appending Rows
Two data frames can be combined with `bind_rows()` **IF** they have the same column names and classes. For example suppose that two other data frames exist -- `bears2` has more information about bears and `bobcats` has similar information about bobcats.

```{r echo=FALSE}
bears2 <- data.frame(length.cm=c(135,142,143),
                     weight.kg=c(100,115,110),
                     loc=rep("Iron",3))
bobcats <- data.frame(length.cm=c(75,82,71,79),
                      weight.kg=c(6.2,8.1,7.4,7.6),
                      loc=c("Douglas","Douglas","Bayfield","Douglas"))
```
```{r}
bears2
bobcats
```

The code below appends the `bears2` data frame to the bottom of the `bears` data frame and then, for demonstration purposes, orders the bears by size.

```{r}
newbears <- bind_rows(bears,bears2) %>%
  arrange(length.cm,weight.kg)
newbears
```

The same could be done with the `bears` and `bobcats` data frames but there will be no way to then tell which observations are for bears and which are for bobcats. This deficiency can be overcome by giving names to the data frames within `bind_rows()` and giving a variable name to `.id=` for the new variable that will identify the groups. For example,

```{r}
animals <- bind_rows("bear"=bears,"bobcat"=bobcats,.id="animal")
animals
```

Note that more than two data frames can be combined with `bind_rows()`.

::: {.tip data-latex=""}
Use `bind_rows()` to combine two (or more) data frames that have the same variables (i.e., columns).
:::

&nbsp;

## Examples in Context
### NBA Players {#rows-nba}
In Section \@ref(columns-nba) the `players2` data frame was created that showed the starting year, ending year, total years played, and whether the player was from the "modern" era or not for all NBA players.

```{r}
players2
```

The graph shown in that same section was for all players with more than 18 years in the NBA. The data frame for that graph is constructed below.

```{r}
nba_gt18 <- players2 %>% filter(years_played>18)
nba_gt18
```

It might be interesting to see who started in the NBA in the year of your college graduation (using mine below).

```{r}
nba_grad1 <- players2 %>% filter(start==1989)
nba_grad1
```

Perhaps those that started in the year of your graduation and played for more than a decade.

```{r}
nba_grad2 <- players2 %>% filter(start==1989,years_played>10)
nba_grad2
```

Perhaps those that were playing during the year of your graduation.

```{r}
nba_grad3 <- players2 %>% filter(start<=1985,end>=1985)
nba_grad3
```

Perhaps those whose name was "Jordan".

```{r}
nba_jordans <- players2 %>% filter(grepl("Jordan",name))
nba_jordans
```

Ahhh ... need to be a little tricky to get all last names that are "Jordan".

```{r}
nba_jordans2 <- players2 %>% filter(grepl("Jordan,",name))
nba_jordans2
```

Also tricky to get all first names that are "Jordan".

```{r}
nba_jordans3 <- players2 %>% filter(grepl(", Jordan",name))
nba_jordans3
```

It is intersting that there were no players in the NBA with the first name "Jordan" before Michael Jordan (the greatest player of all time) retired.

&nbsp;

### Wolves and Moose of Isle Royale {#rows-isle-royale}
In Section \@ref(columns-isle-royale) the `irmw2` data frame was created for use in the graphing course.

```{r}
irmw2
```

One of the things we did in that class is focus on the "early" years of the wolf-moose time series. Such a data frame is create below.

```{r}
irmw_early <- irmw2 %>% filter(era=="early")
```

Other things that we could do with these data are ...

* Find years where the moose population was more than 1500 animals.

```{r}
tmp <- irmw2 %>% filter(moose>1500)
tmp
```

* Find years where the wolf population was less than 10 animals.

```{r}
tmp <- irmw2 %>% filter(wolves<10)
tmp
```

* Find years where the wolf population was less than 10 animals and the moose population was greater than 1500 animals.

```{r}
tmp <- irmw2 %>% filter(wolves<10,moose>1500)
tmp
```

* Find years where ice bridges formed.

```{r}
tmp <- irmw2 %>% filter(ice_bridges=="yes")
tmp
```

* Find years in the 1970s decade and order by the descending number of wolves.

```{r}
tmp <- irmw2 %>%
  filter(year>=1970,year<1980) %>%
  arrange(desc(wolves))
tmp
```

* Find years in the 1980s decade, order by the descending number of wolves, and show only the top three years (i.e., most wolves).

```{r}
tmp <- irmw2 %>%
  filter(year>=1980,year<1990) %>%
  arrange(desc(wolves)) %>%
  slice_head(n=3)
tmp
```