# Factors
Factors are a special case of character variables in R. Factors are primarily used for categorical data, especially if summaries of that data are to be displayed in other than alphabetical order. For example, a simple count of the number of observations in each `loc` of the `bears` data frame introduced in Section \@ref(csv-files) shows that the categories of the character variable are displayed alphabetically by default.

```{r}
bears <- read_csv(file.path("data","Bears.csv"))
tmp <- bears %>%
  group_by(loc) %>%
  summarize(n=n())
tmp
```

To display these results in other than alphabetical order will require the use of factors, which will be illustrated in this module.

Factors have a long and, for some, dubious history in R. The `forcats` package was developed to aid working with factors in R. While it is not critical to understanding the methods of this module, the [stringsAsFactors: An unauthorized biography](http://simplystatistics.org/2015/07/24/stringsasfactors-an-unauthorized-biography/) and [stringsAsFactors = \<sigh\>](http://notstatschat.tumblr.com/post/124987394001/stringsasfactors-sigh) are interesting reads about how unforeseen issues can arise when using factors. Many of these issues are addressed when using `forcats`, which is loaded with `tidyverse`.

&nbsp;

## Creating Factors
A variable is converted to a factor class by including that variable in `factor()`. The code below creates a new data frame where `loc` is now a factor, as indicated by `<fct>` below `loc` when the tibble is printed.

```{r}
tmp <- bears %>%
  mutate(loc=factor(loc))
tmp
```

Nothing is changed when the data are displayed as above. However, behind-the-scenes the factor variable consists of numerical codes that are mapped to levels (i.e., the categories) of the factor variable. The order here is still alphabetical because the order of the levels was not specifically controlled when creating the factor variable. Thus, behind-the-scenes "1" is mapped to "Ashland", "2" to "Bayfield", and "3" to "Douglas". This mapping is shown below.

```{r}
tmp %>% mutate(loc_nums=as.numeric(loc))
```

Suppose, however, that we prefer that the levels of the `loc` variable be ordered from West to East, as in Douglas than Bayfield and then Ashland. The specific order of the levels can be set by supplying the ordered levels in a vector given to `levels=` within `factor()`.

```{r}
tmp <- bears %>%
  mutate(loc=factor(loc,levels=c("Douglas","Bayfield","Ashland")))
tmp
```

Again, the data are not fundamentally altered, but the codes underlying the factor variables are.^[See that "1" is now mapped to "Douglas", "2" to "Bayfield", and "3" to "Ashland".]

```{r}
tmp %>% mutate(loc_nums=as.numeric(loc))
```

When setting the levels of a factor variable make sure that all levels are spelled exactly as they are in the original variable. For example, see the major problem created below when a level is erroneously set (i.e., "bayfield" is not capitalized).

```{r}
tmp <- bears %>%
  mutate(loc=factor(loc,levels=c("Douglas","bayfield","Ashland")))
tmp
```

Additionally, make sure that you include all possible levels when creating the levels.

```{r}
tmp <- bears %>%
  mutate(loc=factor(loc,levels=c("Douglas","Ashland")))
tmp
```

Given these last two issues it is important to understand what levels exist within a variable. If the variable is a character class (and not yet a factor) then use `unique()` to see the list of levels.

```{r}
unique(bears$loc)
```

However, if the variable is a factor class already then use `levels()` to see the list of levels along with their order.

```{r}
levels(tmp$loc)
```

&nbsp;

## Changing Factor Order
Factor levels can be manually reordered using `levels=` as shown in the previous section. However functions in `forcats` provide other methods to simplify reordering levels for common situations.

Levels can be reordered based on the value of another variable with `fct_reorder()`. This function takes the factor variable as the first argument and a numeric variable that contains values for which to reorder the factor levels as the second argument. Optionally a third argument called `.fun` can be used to calculate a summary of the numeric variable to use for reordering.^[Note that `.fun` is only useful if multiple observations of the factor variable are present and it defaults to using the median.]

For example, the code below orders the levels in `loc` by the median length of the bears.

```{r}
tmp <- bears %>%
  mutate(loc=fct_reorder(loc,length.cm))
levels(tmp$loc)

tmp %>%
  group_by(loc) %>%
  summarize(n=n(),
            mdn=median(length.cm))
```

Alternatively, the code below orders the levels of `loc` by the minimum length of the bears.

```{r}
tmp <- bears %>%
  mutate(loc=fct_reorder(loc,length.cm,.fun=min))
levels(tmp$loc)

tmp %>%
  group_by(loc) %>%
  summarize(n=n(),
            min=min(length.cm))
```

&nbsp;

Specific levels are moved to the beginning of the order of levels with `fct_relevel()`. The first argument to this function is the factor variable and subsequent arguments are levels to move to the "beginning of the line." For example, the following code moves "Bayfield" to the beginning of the order.

```{r}
tmp <- bears %>%
  mutate(loc=fct_relevel(loc,"Bayfield"))
levels(tmp$loc)
```

Alternatively, both "Bayfield" and "Douglas" are moved to the beginning below

```{r}
tmp <- bears %>%
  mutate(loc=fct_relevel(loc,"Bayfield","Douglas"))
levels(tmp$loc)
```

&nbsp;

The order of levels for a factor variable are reversed with `fct_rev()`.

```{r}
tmp <- bears %>%
  mutate(loc=fct_rev(loc))
levels(tmp$loc)
```

The `fct_rev()` function can be used with `fct_reorder()` to change the order from ascending to descending order. For example, the code below show the factor levels in descending order of the minimum length.

```{r}
tmp <- bears %>%
  mutate(loc=fct_reorder(loc,length.cm,.fun=min),
         loc=fct_rev(loc))

tmp %>%
  group_by(loc) %>%
  summarize(n=n(),
            min=min(length.cm))
```

&nbsp;

The `fct_infreq()` function is used to order the levels by **de**creasing frequency of their occurrence.

```{r}
tmp <- bears %>%
  mutate(loc=fct_infreq(loc))

tmp %>%
  group_by(loc) %>%
  summarize(n=n())
```

Of course, using `fct_rev()` with `fct_infreq()` would order the levels in ascending frequency of occurrence.

```{r}
tmp <- bears %>%
  mutate(loc=fct_rev(fct_infreq(loc)))

tmp %>%
  group_by(loc) %>%
  summarize(n=n())
```



## Changing Factor Levels
In some instances it may be beneficial to change the names of the levels or to collapse or lump levels together. Methods for performing these changes are demonstrated in this section using data on the density and basal area of tree species located on plots in the Apostle Islands that were designated as "Balsam Fir" plots. These data were extracted from [Sanders and Grochowski (2012)](https://irma.nps.gov/DataStore/DownloadFile/454366)^[Note that Jessica Grochowski, now Jessica Kirschbaum, is a Northland alum.] and are stored in [APIS_FirPlots.xlsx](https://github.com/droglenc/BookWrangling/raw/main/data/APIS_FirPlots.xlsx).

```{r}
aip <- readxl::read_excel(file.path("data","APIS_FirPlots.xlsx"))
aip
```

These data are read in and briefly summarized below.

```{r}
aip_sum1 <- aip %>%
  group_by(species) %>%
  summarize(n=n(),
            ttl_density=sum(density),
            ttl_barea=sum(basal_area))
aip_sum1
```

The summary above shows several species within a few genera. A research may wish to collapse these species into a one level that represents the genera. For example, one may want to collapse "Acer rubrum", "Acer saccharum", "Acer spicatum", and "Acer sp." into a single level of "Acer sp."

The "collapsing" of multiple levels into one level is accomplished with `fct_collapse()`. The first argument to this function is the factor variable containing the "old" levels. Subsequent arguments are formed by setting a new level name equal to a vector containing old level names to collapse.

For example the code below uses `fct_collapse()` to create a new `species2` variable that collapses the multiple "Acer" and "Betula" levels into two levels specific to each genera. The levels of `species2` are then ordered in descending order of the sum of `density` (using `fct_reorder()` and `fct_rev()`).^[Compare the `species` to `species2` columns below to confirm how the "refactoring" worked.]

```{r}
aip %<>%
  mutate(species2=fct_collapse(species,
                               "Acer sp."=c("Acer rubrum","Acer saccharum",
                                            "Acer sp.","Acer spicatum"),
                               "Betula sp."=c("Betula alleghaniensis",
                                              "Betula papyrifera",
                                              "Betula sp."),
                               "Unknown"=c("unknown tree - hardwood",
                                           "unknown tree - softwood")),
         species2=fct_rev(fct_reorder(species2,density,.fun=sum)))
aip
```

Total density and total basal area by level of `species2`, where the levels of `species2` are ordered in descending order of the total density is shown below.

```{r}
aip_sum2 <- aip %>%
  group_by(species2) %>%
  summarize(n=n(),
            ttl_density=sum(density),
            ttl_barea=sum(basal_area))
aip_sum2
```

The result above indicate that four of the "species" dominated the results. It may be useful (especially if graphing these results) to "lump" the four infrequent species into an "Other" category. This type of "lumping" can be accomplished with `fct_lump_n()` which requires the factor variable as its first argument and`n=` to indicate the top `n` levels to **not** lump into an "Other" category. For example, the code below will lump all levels in `species2` after the top four into an "Other" category of the new `species3` variable. The summary results illustrate the lumping.

```{r}
aip <- aip %>%
  mutate(species3=fct_lump_n(species2,n=4))

aip_sum3 <- aip %>%
  group_by(species3) %>%
  summarize(n=n(),
            ttl_density=sum(density),
            ttl_barea=sum(basal_area))
aip_sum3
```

Other versions of `fct_lump_XXX()`, such as `fct_lump_prop()` and `fct_lump_min()`, can be used when ordering by frequency rather than be another variable as shown here.

Finally, the names of specific levels can be changed with `fct_recode()`. The first argument to this function is the original factor variable. Subsequent arguments are of the form new level name equal to old level name. Any levels not listed in `fct_recode()` will be retained with their original names. For example, the code below creates a new `species4` variable with new common names for the three "species" levels remaining in `species3`. The results are seen in the summary below.

```{r}
aip %<>%
  mutate(species4=fct_recode(species3,
                             "Maple" = "Acer sp.",
                             "Birch" = "Betula sp.",
                             "Balsam Fir" = "Abies balsamea"))

aip_sum4 <- aip %>%
  group_by(species4) %>%
  summarize(n=n(),
            ttl_density=sum(density),
            ttl_barea=sum(basal_area))
aip_sum4
```




## Examples in Context
### Pulse of the Nation Survey
[Cards Against Humanity Saves America](https://www.cardsagainsthumanitysavesamerica.com/) created monthly polls to assess what they called the "Pulse of the Nation." "Cards" partnered with Survey Sampling International — a professional research firm — to contact a nationally representative sample of the American public. For the first three polls, they interrupted people’s dinners on both their cell phones and landlines, and a total of about 3000 adults did not hang up immediately. Results from the first poll are in [201709-CAH_PulseOfTheNation.csv]()^[These data are from [Kaggle](https://www.kaggle.com/cardsagainsthumanity/pulse-of-the-nation)], which are read in below.

```{r}
PON <- read_csv(file.path("data","201709-CAH_PulseOfTheNation.csv"))
PON
```

The variables in the data frame are generally the actual questions asked in the survey. These are both difficult to read and to work with. To get a better look at these variables I used `names()`, which when given a data frame returns all the variable names of that data frame.

```{r}
names(PON)
```

For this analysis I want to focus on the `Political Affiliation`, `Age Range`, and the three variable related to education, opinion about climate change, and opinion about the honesty of scientists. These last three variables have very long names so it will be easier to refer to them by their column **number**. From above, it is noted that the "education" variable is column 7, the climate change variable is in column 15, and the "scientists honesty" question is in column 18. In the code below, I rename each of the columns (and `Age Range`) and then reduce the data frame to just those variables.

```{r}
PON %<>%
  rename(polit_aff=`Political Affiliation`,age_range=`Age Range`,
         education=7,climate=15,scihon=18) %>%
  select(polit_aff,age_range,education,climate,scihon)
PON
```

All remaining variables are categorical and for my analysis purposes I want each to be a factor. Before converting these variables to factors I will examine the levels that appear in each.

```{r}
unique(PON$polit_aff)
unique(PON$age_range)
unique(PON$education)
unique(PON$climate)
unique(PON$scihon)
```

From this examination of levels, I made the following notes:

* Respondents were allowed to answer "Don't Know" or refuse to answer which were coded as "DK/REF". My general preference is to have "DK/REF" answers be the last level.
* The levels for `age_range` will **not** need to be controlled because the numbers will naturally alphabetically order (as the numbers are all two digits; the order might not be natural if there were some one digit (say "1-17") or three digit (say "100+") numbers).
* Levels in `education` and `scihon` need to be ordered according to their natural order (i.e., increasing education and increasing level of agreement).
* I will order the levels for `climate` based on my understanding of what the correct answer is (i.e., I will order the answers from correct to less correct).
* I will order the levels of `polit_aff` so that "Independent" is in the middle (which will happen alphabetically) and "DK/REF" is last.

Creation of these factors is below. Note that to move "DK/REF" to the end of `polit_aff` I used `fct_relevel()` with `after=Inf` to move the level to the last position.

```{r}
PON %<>%
  mutate(polit_aff=factor(polit_aff),
         polit_aff=fct_relevel(polit_aff,"DK/REF",after=Inf),
         age_range=factor(age_range),
         education=factor(education,levels=c("High school","Some college",
                                             "College degree","Graduate degree",
                                             "Other","DK/REF")),
         climate=factor(climate,levels=c("Real and Caused by People",
                                         "Real but not Caused by People",
                                         "Not Real At All",
                                         "DK/REF")),
         scihon=factor(scihon,levels=c("Strongly Disagree","Somewhat Disagree",
                                       "Neither Agree nor Disagree",
                                       "Somewhat Agree","Strongly Agree",
                                       "DK/REF")))
```

I then used `levels()` to double-check that each factor has the levels and the order of levels that I expected.

```{r}
levels(PON$polit_aff)
levels(PON$age_range)
levels(PON$education)
levels(PON$climate)
levels(PON$scihon)
```

Suppose the first analysis of interest is examining responses to the "scientists are generally honest and are serving the public good" question.

```{r}
sci_sum1 <- PON %>%
  group_by(scihon) %>%
  summarize(freq=n())
sci_sum1
```

Suppose that the analysts want to simplify this result for their audience by ignoring the "Don't know or refused to answer" responses and collapsing all disagree answers to one level and all agree answers to one level.

```{r}
tmp <- PON %>%
  filter(scihon!="DK/REF") %>%
  mutate(scihon=fct_collapse(scihon,
                             "Disagree"=c("Strongly Disagree","Somewhat Disagree"),
                             "Agree"=c("Strongly Agree","Somewhat Agree")))
sci_sum2 <- tmp %>%
  group_by(scihon) %>%
  summarize(freq=n())
sci_sum2
```

For fun, a graph of this summary is below.

```{r echo=FALSE, fig.width=4.5}
ggplot(sci_sum2,mapping=aes(x=scihon,y=freq)) +
  geom_bar(stat="identity",color="black",fill="gray70") +
  scale_y_continuous(name="Frequency of Responsdents",
                     expand=expansion(mult=c(0,0.05))) +
  labs(title='Agreement with "scientists are generally honest\nand are serving the public good"',
       caption="Data from Cards Against Humanity Saves America") +
  theme_bw() +
  theme(axis.title.x=element_blank())
```

Further suppose that the analysts want to summarize the question about climate change separately by political affiliation (for which an answer was provided).

```{r}
polclim_sum1 <- PON %>%
  filter(polit_aff != "DK/REF") %>%
  group_by(polit_aff,climate) %>%
  summarize(freq=n()) %>%
  mutate(perc=freq/sum(freq)*100)
polclim_sum1
```

Suppose that the researchers want to compute a percentage of respondents that thought climate change was real, whether they had the cause correct or not. Further suppose that they wish to include the "DK/REF" answers with the "Not Real At All" answers under the assumption that if they though climate change was real they would have responded as such.

```{r}
tmp <- PON %>%
  filter(polit_aff != "DK/REF") %>%
  mutate(climate=fct_collapse(climate,
                              "Real"=c("Real and Caused by People",
                                       "Real but not Caused by People"),
                              "Not Real"=c("Not Real At All",
                                           "DK/REF")))
polclim_sum2 <- tmp %>%
  group_by(polit_aff,climate) %>%
  summarize(freq=n()) %>%
  mutate(perc=freq/sum(freq)*100) %>%
  ungroup()
polclim_sum2
```

And a graphic of the results for fun.

```{r echo=FALSE, fig.width=4.5}
ggplot(polclim_sum2,mapping=aes(x=polit_aff,y=perc,fill=fct_rev(climate))) +
  geom_bar(stat="identity") +
  scale_y_continuous(name="Frequency of Responsdents",
                     expand=expansion(mult=0)) +
  scale_fill_manual(values=c("Real"="gray30","Not Real"="gray70"),guide="none") +
  annotate(geom="text",x=1,y=50,label="Real",color="white") +
  annotate(geom="text",x=3,y=85,label="Not Real") +
  labs(title='Response about Climate Change being Real',
       subtitle='by Political Affiliation',
       caption="Data from Cards Against Humanity Saves America") +
  theme_bw() +
  theme(axis.title.x=element_blank())
```

&nbsp;

### Enterococci on Sydney Beaches
The Beachwatch Water Quality Program of Sydney, Australia measured the density (in colony forming units per 100 ml of water) of [*Enterococci*](https://en.wikipedia.org/wiki/Enterococcus) on Sydney beaches. The data from 2013 to 2018 was made available by the R-Ladies Sydney group and is in [sydneybeaches.csv](https://raw.githubusercontent.com/rladiessydney/RYouWithMe/master/sydneybeaches.csv).^[Data from [this online book](https://rladiessydney.org/courses/ryouwithme/01-basicbasics-3/).] The data are loaded below.

```{r}
sb <- read_csv("https://raw.githubusercontent.com/rladiessydney/RYouWithMe/master/sydneybeaches.csv")
sb
```

For the demonstration here the `BeachId`, `Council`, and `Region` variables are not needed. The density of Enterococci is the last variable but its name is difficult to work with and will be renamed.

```{r}
sb %<>%
  select(-BeachId,-Council,-Region) %>%
  rename(density=last_col())
sb
```

In one part of their analysis the researchers wanted to produce a plot of the mean and range of densities of *Enterococci* by beach. To aid understand, however, they wanted the beaches ordered from south to north. The summary statistics below are calculated by `Site` converted to a factor class based on `Latitude`. The latitudes are list with negative numbers (because Australia is in the southern hemisphere) so that more negative numbers are further south.

```{r}
sb_sum1 <- sb %>%
  mutate(Site=fct_reorder(Site,Latitude)) %>%
  group_by(Longitude,Latitude,Site) %>%
  summarize(n=n(),
            valid_n=sum(!is.na(density)),
            mn_density=mean(density,na.rm=TRUE),
            sd_density=sd(density,na.rm=TRUE),
            min_density=min(density,na.rm=TRUE),
            max_density=max(density,na.rm=TRUE)) %>%
  ungroup()
sb_sum1
```

The desired plot is shown below (with the southern-most beach at the bottom of the y-axis).

```{r echo=FALSE, fig.width=5}
sb_sum1 %<>%
  mutate(Site=fct_rev(Site))
sb_plot1 <- ggplot(sb_sum1) +
  geom_point(mapping=aes(x=mn_density,y=Site)) +
  geom_segment(mapping=aes(x=min_density,xend=max_density,
                           y=Site,yend=Site),color="gray60") +
  scale_x_continuous(name="Density of Enterococci (cfu/100 ml)",
                     expand=expansion(mult=0.01)) +
  labs(title="Enterococci Density on Sydney Beaches",
       subtitle="Mean and range for 2013-2018 shown",
       caption="Data from R-Ladies Sydney") +
  theme_bw() +
  theme(axis.title.y=element_blank())
sb_plot1
```

Other researchers would prefer that the beaches are ordered from lowest to highest density of *Enterococci*. The code below creates a summary by `Site`s ordered by the mean `density`. Note the use of `na.rm=TRUE` in `fct_reorder()` because of the missing values at some sites.

```{r}
sb_sum2 <- sb %>%
  mutate(Site=fct_reorder(Site,density,.fun=mean,na.rm=TRUE)) %>%
  group_by(Site,Longitude,Latitude) %>%
  summarize(n=n(),
            valid_n=sum(!is.na(density)),
            mn_density=mean(density,na.rm=TRUE),
            sd_density=sd(density,na.rm=TRUE),
            min_density=min(density,na.rm=TRUE),
            max_density=max(density,na.rm=TRUE)) %>%
  ungroup()
sb_sum2
```

```{r echo=FALSE, fig.width=5}
sb_plot1 %+% mutate(sb_sum2,Site=fct_rev(Site))
```
